### 2023-03-22 chapter0/chapter0.tex
Added more citations covering FlyBase, conformational classes, disorder prediction, and AD sequence features

Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues. However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues. Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{vanderLee2014, Das2015}. For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other. However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational class depends on the distribution of those charged residues. If the two classes of residues are segregated into separate blocks of like charges, they attract and form hairpins. However, if they are evenly distributed, the attractive and repulsive forces balance on average, and the sequence generally assumes expanded coil-like conformations. High numbers of polar amino acids, which are hydrophilic but not charged, are associated with semi-compact globules. Though their side chains, the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic backbone, composed the donor and receptor sites common to all amino acids, to self-associate. Thus, like folded domains, the structures of disordered proteins are dictated by their sequences. However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational class.

@@

Both transcription and translation are tightly regulated processes, but because cells employ numerous interlocking mechanisms to control access to DNA and the creation of mRNA, transcription is often the major regulatory checkpoint in gene expression. A key step in this process is the recruitment of the protein complex which transcribes DNA into mRNA to the beginning of a gene's sequence, \textit{i.e.} its transcription start site (TSS). The placement of this complex, called RNA polymerase, is dictated by regulatory sequences encoded in the DNA, and while some regulatory sequences are ubiquitous, like those that mark a gene's TSS, many are gene-specific. Thus, the unique collection of regulatory sequences associated with a gene determines many aspects of its expression. RNA polymerase does not directly contact these gene-specific regulatory elements, however. Instead, proteins called transcription factors directly bind these sequences and interact with other components of the transcriptional machinery which in turn recruit RNA polymerase to the TSS. Accordingly, a typical transcription factor has two parts: a DNA-binding domain and an activation domain. DNA-binding domains are usually structured because they make stable contacts with specific regulatory sequences. In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs~\cite{Liu2006}. Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Staller2018, Sanborn2021}. Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology. As disordered regions play a key role in this process, studies which broadly investigate the relationship between the sequence, structure, and function of IDRs therefore have significant implications for our understanding of gene regulation.

@@

Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone, the relationship between the sequence and function remains poorly understood. In contrast, predictions of structure and function are readily available for many folded domains. Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly. Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}. Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades. Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences. Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences. Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.

@@

These analyses were conducted using IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research. However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems. As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021}. Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease. The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution. For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast. In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.

>>>>

Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues~\cite{Uversky2000}. However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues. Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{Mao2010, vanderLee2014, Das2015}. For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other. However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational class depends on the distribution of those charged residues. If the two classes of residues are segregated into separate blocks of like charges, they attract and form hairpins. However, if they are evenly distributed, the attractive and repulsive forces balance on average, and the sequence generally assumes expanded coil-like conformations. High numbers of polar amino acids, which are hydrophilic but not charged, are associated with semi-compact globules. Though their side chains, the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic backbone, composed the donor and receptor sites common to all amino acids, to self-associate. Thus, like folded domains, the structures of disordered proteins are dictated by their sequences. However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational class.

@@

Both transcription and translation are tightly regulated processes, but because cells employ numerous interlocking mechanisms to control access to DNA and the creation of mRNA, transcription is often the major regulatory checkpoint in gene expression. A key step in this process is the recruitment of the protein complex which transcribes DNA into mRNA to the beginning of a gene's sequence, \textit{i.e.} its transcription start site (TSS). The placement of this complex, called RNA polymerase, is dictated by regulatory sequences encoded in the DNA, and while some regulatory sequences are ubiquitous, like those that mark a gene's TSS, many are gene-specific. Thus, the unique collection of regulatory sequences associated with a gene determines many aspects of its expression. RNA polymerase does not directly contact these gene-specific regulatory elements, however. Instead, proteins called transcription factors directly bind these sequences and interact with other components of the transcriptional machinery which in turn recruit RNA polymerase to the TSS. Accordingly, a typical transcription factor has two parts: a DNA-binding domain and an activation domain. DNA-binding domains are usually structured because they make stable contacts with specific regulatory sequences. In contrast, activations domains interact with the transcriptional machinery and are highly enriched in IDRs~\cite{Liu2006}. Decades of research have identified many of the proteins involved in these interactions as well as some common properties of their sequences~\cite{Ma1987, Sigler1988, Gerber1994, Arnold2018, Staller2018, Erijman2020, Sanborn2021}. Despite this progress, however, the precise mechanisms by which transcription factors locate their DNA targets and recruit the transcriptional machinery to initiate expression of specific genes remains an unsolved problem in molecular biology. As disordered regions play a key role in this process, studies which broadly investigate the relationship between the sequence, structure, and function of IDRs therefore have significant implications for our understanding of gene regulation.

@@

Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone, the relationship between the sequence and function remains poorly understood~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}. In contrast, predictions of structure and function are readily available for many folded domains. Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly. Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}. Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades. Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences. Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences. Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.

@@

These analyses were conducted using IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research. However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems. As another foundational model organism with abundant genomic information across many evolutionary lineages, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}. Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in human health and disease. The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution. For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast. In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
